Individual Partisan Harm and Utility in Redistricting

### 2021-10-14 paper/paper.Rmd
fix appendices
small edits

---
title: Individual Partisan Harm and Utility in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed. 
      The authors thank {you} for helpful comments and discussion.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Most existing standards for assessing partisan gerrymandering focus on the symmetric treatment of parties by electoral systems. We develop measures of utility and harm for individual voters by considering counterfactual redistricting plans, and use these notions to construct a normative standard of partisan fairness which is centered on the individual. We estimate these measures from real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements. We apply our measures to congressional redistricting in New Jersey, and compare them to existing measures of partisan symmetry and competitiveness.
keywords:
  - partisan fairness
  - representation
  - redistricting simulation
  - individual standing
bibliography: references.bib
biblio-style: apalike
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: sentence
---

@@

We argue that understanding the redistricting process requires a feasible and reliable understanding of a counterfactual set of redistricting plans.
To that end, we define two individual utility measures for districting plans: a *local utility* capturing the value to a voter of having a representative of the same party, and a *global utility* reflecting the value in maximizing the number of seats awarded to a voter's chosen party.
From these two utility measures, we derive a measure of harm, based on counterfactual plans, and a notion of individual partisan fairness, which occurs when harms are evenly distributed across voters of each party.
Deviations from fairness are measured by *differential harm*, which, as we show, is intimately connected to the number of seats won by each party.

@@

To estimate these measures requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
Early methodologies for simulating plans relied on randomness to generate many different districting plans [@ciri:darl:orou:00; @chen:rodd:13; @deford2019]. These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations. Yet, lacking a specific target probability distribution, these heuristic approaches to simulation do not allow for probabilistic interpretations, and risk misrepresenting the total space of possible districting plans [@fifield2020essential]. Recent advances in Monte Carlo methods allow us move beyond random simulation and towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020]. We can thus develop a reliable set of counterfactual plans that can be tuned to reflect the constraints imposed by the laws and practices of a particular state [@matt:hers:21].

@@

```{=tex}
\clearpage
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```
# Proofs of Propositions {#app:proofs}

@@

\begin{proof}
\end{proof}
```
```{=html}
<!--
# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}

>>>>

---
title: Individual Partisan Harm and Utility in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed. 
      The authors thank {you} for helpful comments and discussion.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Most existing standards for assessing partisan gerrymandering focus on the symmetric treatment of parties by electoral systems. We develop measures of utility and harm for individual voters by considering counterfactual redistricting plans, and use these notions to construct a normative standard of partisan fairness which is centered on the individual. We calculate these measures on real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements. We apply our measures to congressional redistricting in New Jersey, and compare them to existing measures of partisan symmetry and competitiveness.
keywords:
  - partisan fairness
  - representation
  - redistricting simulation
  - individual standing
bibliography: references.bib
biblio-style: apalike
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: sentence
---

@@

We argue that understanding the redistricting process requires a counterfactual set of feasible redistricting plans.
To that end, we define two individual utility measures for districting plans: a *local utility* capturing the value to a voter of having a representative of the same party, and a *global utility* reflecting the value in maximizing the number of seats awarded to a voter's chosen party.
From these two utility measures, we derive a measure of harm, based on counterfactual plans, and a notion of individual partisan fairness, which occurs when harms are evenly distributed across voters of each party.
Deviations from fairness are measured by *differential harm*, which, as we show, is intimately connected to the number of seats won by each party.

@@

To calculate these measures in practice requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
Early methodologies for simulating plans relied on randomness to generate many different districting plans [@ciri:darl:orou:00; @chen:rodd:13; @deford2019]. These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations. Yet, lacking a specific target probability distribution, these heuristic approaches to simulation do not allow for probabilistic interpretations, and risk misrepresenting the total space of possible districting plans [@fifield2020essential]. Recent advances in Monte Carlo methods allow us move beyond random simulation and towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020]. We can thus develop a reliable set of counterfactual plans that can be tuned to reflect the constraints imposed by the laws and practices of a particular state [@matt:hers:21].

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```
# Proofs of Propositions {#app:proofs}

@@

\begin{proof}
\end{proof}
```
# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}

```{=html}
<!--
